Profitability Prediction of Consulting Projects in the Construction Industry

### 2016-08-31 finalwriteup/draft.rmd
paul mark ups

Engaging in loss making jobs for fixed fees is a major problem in consulting, particularly in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help avoid loss making jobs or help tactfully choose when to enforce stricter contracts. It was found that in a specific decision framework, a case study's profits could be improved 9% by avoiding approximately 4% of projects. Alternative decision frameworks are also proposed and evaluated. The internal data collected by the case study company described each of their projects in terms of employee positions, fraction of project time completed by each employee, client characteristics, project characteristics (time span, hours, technical details) and invoicing history. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naive Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario that rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. This process was repeated on hundreds of models built from different training subsets of the data in order to obtain a 95% statistical confidence interval of profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. Further work is recommended to test user input estimation and user interface designs. The findings from this research have the potential to assist managers in reducing losses by highlighting risky projects and guiding project-based changes to fee structures.

@@

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the task is new and difficult. These errors are trivial in everyday activities, but for consulting businesses that solve complex problems, similar time estimation errors determine their financial well being.

Consulting businesses give expert advice to other professionals in exchange for a fee based on the amount of time their consulting staff spends on the project. The deliverable 'expert advice' is generally in the form of a document or a package of documents. A document could range from a 5-page report written by a single person to a sequence of hundreds of drawings assembled by a team of 10 over many years. Despite the range of possible deliverables, before any project commences, the client and consultant must agree on a fee, or at least a fee structure. The fee structure can be as creative as the engaged parties wish, but in fact tend to fit a few common models. Two examples are:

@@

In each of these contract models, financial risk is assigned to one of the parties. In model 1, the client must pay for an unknown quantity of consultants' billable hours and in model 2 the consultants must complete the project for a fixed price while not knowing how much of their own time will be spent.

@@

The way a consulting manager calculates a fixed price varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and, after reviewing the project details, can use a combination of intuition, comparison to past projects, and rules of thumb (such as a set fraction of the entire client budget for a large project). Such qualitative methods can result in profit margins, which ultimately depend on the skill (or luck) of the manager in predicting the amount of time a project demands.

This thesis focuses on a single case study consulting company in the construction industry that has persevered through many fixed price projects that resulted in losses. A source of insight into the loss making projects was identified - twelve years of passively collected data that described each project in terms of their clients, invoice history, employee hours and technical details. Was it possible to find trends in what makes a project loss making and can these characteristics be identified before a fixed price is offered? Can these trends be modeled into a predictive algorithm, and if so, how could the outcomes become integrated into business decisions? Finally, if the predictive outcomes were integrated into decision-making, would they definitely improve the business' bottom line? These questions are explored in depth and answered over the course of this dissertation.

@@

The three methods were anticipated to output different levels of influence for each variable. The tested methods were ANOVA Linear Regression, conditional inference forests (cForests), and Random Forests. It was expected that Random Forests would favour variables that had more categories, a well-known bias; which is also a bias that cForests have overcome [@Strobl2007]. ANOVA is based on linear theory, which contrasts conditional forest and Random Forest's more free-form structure. Therefore, it is beneficial to compare the methods' assessments of variable importance.

@@

Six blending methods, ranging from simple to complex, were tested using predictions from the top performing individual models. These included simple averaging of the individual model results, building a Logistic Regression model using the individual model results only, a Boosted Tree model using individual model results only, feature weighted linear stacking (FWLS), Random Forests, and Boosted Trees. Where possible, these methods shall be expressed as equations in an example problem with three original explanatory variables, $X_1$, $X_2$, $X_3$, and two full sets of predictions from individual predictive models, $M_1$ and $M_2$, with a response variable $Y$. Values in $M_1$ and $M_2$ are probabilities between 0 and 1.

@@

$Y = \frac{1}{2} \cdot (M_1 + M_2)$

@@

$Y = \beta_0 + \beta_1 M_1 + \beta_2 M_2$

The third method builds Boosted Trees from the individual models, i.e. using $M_1$ and $M_2$ only to predict $Y$ in a Boosted Tree model. The last three methods take advantage of potentially meaningful interactions between the individual models' predictions and meta-features. For example, if the $M_1$ model predicted profitability better than $M_2$ for projects in category $a$ of $X_1$, the blending model would take advantage of this interaction. $M_1$ would be weighted higher than $M_2$ for projects where $X_1 = a$.

@@

$$
\begin{aligned} 
Y &= \beta_0 + \beta_1 M_1 + \beta_2 M_2 + \beta_3 X_1 + \beta_4 X_2 + \beta_5 X_3 + \\
\qquad \beta_6(M_1 \cdot X_1) + \beta_7(M_1 \cdot X_2) + \beta_8(M_1 \cdot X_3) +\\
\qquad \beta_9(M_2 \cdot X_1) + \beta_10(M_2 \cdot X_2) + \beta_11(M_2 \cdot X_3)
\end{aligned}
$$

Random Forests and Boosted Trees were also used to blend meta-features with output from the best models. Feature interaction is performed passively due to the nature of how trees are built. A split in a node determined by one variable is conditional upon the preceding split, which was based on another variable and so on. Random Forests and Boosted Tree models are difficult to express as equations, but for the example problem, $X_1$, $X_2$, $X_3$, $M_1$, and $M_2$ would all be included as explanatory variables to predict $Y$.

@@

$RMSE_b = \sqrt{\frac{\sum_{t = 1}^{n} (\widehat{y}_t - \bar y)^2}{n}}$

@@

$$
\begin{aligned}
RMSE_b &= \textrm{baseline root mean squared error} \\
n &= \textrm{number of data points} \\
\widehat{y} &= \textrm{predicted values of the response variable} \\
\bar y &= \textrm{mean 'return per dollar' for all projects} \\
\end{aligned}
$$

@@

If the models were effective, the RMSE should be lower for ANOVA model predictions, so the difference would be greater than 0. Below is a histogram of this difference across 50 models run on randomly sampled 75% train and 25% test sets using core variables only (method 1 ANOVA). Note, only `r n.samples(data = results.A2$diff)` samples (models) were required for a statistical power of 80% [@pwr].

@@

It was finally decided that Random Forest and Boosted Trees would likely not be distinguishable since the power calculation indicated that 700 models were required. Hence, 100 models of each method were built as calculated for Logistic Regression. This was achieved by performing 10-fold cross validation 10 times. The violin plot below summarises the AUC values produced by each method and the 'violins' are coloured according to whether the distributions significantly vary to Boosted Trees using a critical value of 0.05.

@@

It is not clear why simple linear models and the averaging model outperformed ensemble tree blending methods. As previously stated, the ensemble trees may not have received enough data to adequately learn the complex series of rules they develop. It was also surprising FWLS was ineffective given the success it had in the Netflix Prize competition [@Sill2009]. The only differences between FWLS and the simple Logistic Regression (that performed best) were two additional explanatory variables and four interaction terms. The number of variables was not high for the amount of data since according to @Peduzzi1996, 10 events per explanatory variable or more avoids the risk of biased estimation of variable coefficients in Logistic Regression. The data contained 315 events per explanatory variable and did not pose that risk. Nevertheless, the additional variables must have added misleading noise to the model.

>>>>

Engaging in loss making jobs for fixed fees is a major problem in consulting, particularly in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help avoid loss making jobs or help tactfully choose when to enforce stricter contracts. It was found that in a model-informed decision framework, a case study's profits could be improved 9% by avoiding approximately 4% of projects. Alternative decision frameworks are also proposed and evaluated. The internal data collected by the case study company described each of their projects in terms of employee positions, fraction of project time completed by each employee, client characteristics, project characteristics (time span, hours, technical details) and invoicing history. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naive Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario that rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. This process was repeated on hundreds of models built from different training subsets of the data in order to obtain a 95% statistical confidence interval of profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% improvement in profit. Further work is recommended to test user input estimation and user interface designs. The findings from this research have the potential to assist managers in reducing losses by highlighting risky projects and guiding project-based changes to fee structures.

@@

Predicting how long it will take a person to do something is a notoriously tricky task. People make mental calculations of this sort every day - estimating how long it will take to drive to meet someone on time, how many tasks you can check off in a day, or how long it will take to cook a dish. As most people have experienced, these estimates can be particularly inaccurate if the task is new and difficult. These errors are trivial in everyday activities, but for consulting businesses that solve complex problems, similar time estimation errors determine their financial wellbeing.

Consulting businesses give expert advice to other professionals in exchange for a fee based on the amount of time their consulting staff spends on the project. The deliverable 'expert advice' is generally in the form of a document or a package of documents. A document could range from a 5-page report written by a single person to a sequence of hundreds of drawings assembled by a team of 10 over many years. Despite the range of possible deliverables, before any project commences, the client and consultant must agree on a fee, or at least a fee structure. The fee structure can be as creative as the engaged parties wish, but in fact tend to fit a few common models [@Malhotra2009]. Two examples are:

@@

In each of these contract models, financial risk is assigned to one of the parties. In model 1, the client must pay for an unknown quantity of consultants' billable hours and in model 2 the consultants must complete the project for a fixed price while not knowing how much of their own time will be spent [@Malhotra2009].

@@

The way a consulting manager calculates a fixed price varies from industry to industry and even company to company. Typically a consulting manager has experience in the type of project they are quoting and, after reviewing the project details, can use a combination of intuition, comparison to past projects, and rules of thumb (such as a set fraction of the entire client budget for a large project). Such qualitative methods can impact profit margins, which ultimately depend on the skill (or luck) of the manager in predicting the amount of time a project demands.

This thesis focuses on a single case study consulting company in the construction industry that has persevered through many fixed price projects that resulted in losses. A source of insight could be identified from twelve years of passively collected data that described each project in terms of their clients, invoice history, employee hours and technical details. Was it possible to find trends in what makes a project loss making and can these characteristics be identified before a fixed price is offered? Can these trends be modeled into a predictive algorithm, and if so, how could the outcomes become integrated into business decisions? Finally, if the predictive outcomes were integrated into decision-making, would they definitely improve the business' bottom line? These questions are explored in depth and answered over the course of this dissertation.

@@

Once outliers were removed, the next step was to determine a succinct list of variables that contributed most to the prediction. The three chosen variable importance methods were anticipated to output different levels of influence for each variable. These methods were ANOVA Linear Regression, conditional inference forests (cForests), and Random Forests. It was expected that Random Forests would favour variables that had more categories, a well-known bias; which is also a bias that cForests have overcome [@Strobl2007]. ANOVA is based on linear theory, which contrasts with conditional forest and Random Forest's more free-form structure. Therefore, it is beneficial to compare the methods' assessments of variable importance.

@@

Six blending methods, ranging from simple to complex, were tested using predictions from the top performing individual models. These included simple averaging of the individual model results, building a Logistic Regression model using the individual model results only, a Boosted Tree model using individual model results only, feature weighted linear stacking (FWLS), Random Forests, and Boosted Trees. Where possible, these methods shall be expressed as equations in an example problem with three original explanatory variables, $X_1$, $X_2$, $X_3$, and two full sets of predictions from individual predictive models, $M_1$ and $M_2$, with a response variable $p (probability)$. Values in $M_1$ and $M_2$ are probabilities between 0 and 1.

@@

$p = \frac{1}{2} \cdot (M_1 + M_2)$

@@

$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 M_1 + \beta_2 M_2$

The third method builds Boosted Trees from the individual models, i.e. using $M_1$ and $M_2$ only to predict $p$ in a Boosted Tree model. The last three methods take advantage of potentially meaningful interactions between the individual models' predictions and meta-features. For example, if the $M_1$ model predicted profitability better than $M_2$ for projects in category $a$ of $X_1$, the blending model would take advantage of this interaction. $M_1$ would be weighted higher than $M_2$ for projects where $X_1 = a$.

@@

$$
\begin{aligned} 
\log\left(\frac{p}{1-p}\right) &= \beta_0 + \beta_1 M_1 + \beta_2 M_2 + \beta_3 X_1 + \beta_4 X_2 + \beta_5 X_3 + \\
\qquad \beta_6(M_1 \cdot X_1) + \beta_7(M_1 \cdot X_2) + \beta_8(M_1 \cdot X_3) +\\
\qquad \beta_9(M_2 \cdot X_1) + \beta_10(M_2 \cdot X_2) + \beta_11(M_2 \cdot X_3)
\end{aligned}
$$

Random Forests and Boosted Trees were also used to blend meta-features with output from the best models. Feature interaction is performed passively due to the nature of how trees are built. A split in a node determined by one variable is conditional upon the preceding split, which was based on another variable and so on. Random Forests and Boosted Tree models are difficult to express as equations, but for the example problem, $X_1$, $X_2$, $X_3$, $M_1$, and $M_2$ would all be included as explanatory variables to predict $p$.

@@

$RMSE_b = \sqrt{\frac{\sum_{t = 1}^{n} (\bar y - y_t)^2}{n}}$

@@

$$
\begin{aligned}
RMSE_b &= \textrm{baseline root mean squared error} \\
n &= \textrm{number of data points} \\
\bar y &= \textrm{mean 'return per dollar' as predicted value of the response variable} \\
y_t &= \textrm{observed 'return per dollar' values} \\
\end{aligned}
$$

@@

If the models were effective, the RMSE should be lower for ANOVA model predictions, so the difference would be greater than 0. Because the size of the data set was limited, different divisions of the data into training and test sets would yield different RMSE statistics from the test set. Therefore, multiple models were built for each method in order to understand the distribution of resulting test statistics. Below is a histogram of $\Delta\ RMSE$ across 50 models run on randomly sampled 75% train and 25% test sets using core variables only (method 1 ANOVA). Note, only `r n.samples(data = results.A2$diff)` samples (models) were required for a statistical power of 80% [@pwr].

@@

It was finally decided that Random Forest and Boosted Trees would likely not be distinguishable since the power calculation indicated that 700 models were required. Hence, 100 models of each method were built as calculated for Logistic Regression. This was achieved by performing 10-fold cross validation 10 times. The violin plot below summarises the distributions of the AUC values, where each violin is built from a pair of rotated kernel density plots of the subset of AUC values. The 'violins' are coloured according to whether the distributions significantly vary to Boosted Trees using a critical value of 0.05.

@@

It is not clear why simple linear models and the averaging model outperformed ensemble tree blending methods. As previously stated, the ensemble trees may not have received enough data to adequately learn the complex series of rules they develop. It was also surprising FWLS was ineffective given the success it had in the Netflix Prize competition [@Sill2009]. The only differences between FWLS and the simple Logistic Regression (that performed best) were two additional explanatory variables and four interaction terms. The number of variables was not high for the amount of data since according to @Peduzzi1996, 10 events per explanatory variable or more avoids the risk of biased estimation of variable coefficients in Logistic Regression. The data contained 315 events per explanatory variable and did not pose that risk. Nevertheless, the additional variables might have added misleading noise to the model.
